A Large and Variable Leading Tail of Helium in a Sub-Saturn Undergoing Runaway Inflation

### 2023-07-05 paper/paper1/ms.tex
Addressing comments from Suvrath

\affil{\PSUAA}
\affil{\PSUCEHW}

@@

Inflated hot Saturns stand out as an especially extreme category that should exhibit mass loss.  This category is defined as having masses comparable to Saturn's ($M_p \sim 0.3 M_\mathrm{Jup}$), with equilibrium temperatures high enough to expect radius inflation($T_\mathrm{eq}>1000\;$K).  Their low gravitational potentials should let go of their atmospheres more readily than their hot Jupiter counterparts.  Their lower gravity implies larger atmospheric scale heights, making them easier to detect in transmission spectroscopy.  And their large transit depths and short periods should make them readily detectable in large numbers, like hot Jupiters.

@@

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Here, lightly thermally ionized atmospheric flows induce drag in a planetary magnetic field, weakly coupling the stellar incident energy into the planetary interior.  A key prediction of Ohmic dissipation is that the anomalous heating efficiency $\epsilon(T_\mathrm{eq})$ should exhibit a peak around $T_\mathrm{eq}\sim1500-2000\;$K \citep{2012ApJ...745..138M,2014ApJ...794..132R,2016ApJ...819..116G}. \citet{2018AJ....155..214T} favored Ohmic dissipation as the mechanism responsible for inflating hot Jupiters by showing that the extent of inflation implies an underlying peak at $\sim1500\;$K.  Therefore Ohmic dissipation stands out as a leading physical mechanism for inflation and mass loss.

@@

A runaway inflation scenario predicts an inevitable and profound mass loss rate for inflated hot Saturns.  As the planet's atmosphere overflows its Roche lobe at an ever-increasing pace, instantaneous mass loss rates may exceed $\dot{M}>10^{13}$ g/s, over 10$\times$ larger than those previously seen in atmospheric escape measurements to date \citep{2022arXiv221116243D}.  Inflated hot Saturns, therefore, make excellent targets for direct measurement of atmospheric escape and to test the underlying mechanisms of mass loss.

@@

The evolutionary state of HAT-P-67~b offers even more intriguing possibilities.  The planet may have undergone re-inflation \citep{2022AJ....163...53S, 2022AJ....163..120G, 2023arXiv230306728G}, as the host star evolved through the subgiant phase-- with the planet's insolation increasing with the star's rapid luminosity increase in this part of the short-lived HR diagram.  The tidal gravity of the nearby massive ($M_\star \sim1.6\;M_\odot$) host star may amplify mass loss rates even further \citep{2007A&A...472..329E, 2023ApJ...945L..36T}. Its status as a rare inflated hot Saturn makes HAT-P-67~b a promising laboratory, uniquely suited for testing the runaway inflationary predictions of Ohmic dissipation and XUV photoevaporation.

@@

The observation schedule was strategically restricted to the spring season when the Barycentric Earth Radial Velocity (BERV) would Doppler shift telluric absorption lines sufficiently far away from the core of the Helium 10833 \AA~ feature \citep{2022ApJ...939L..11S}.  The small BERV still means the redward Helium line wing has significant telluric contamination between 10834$-$10836 \AA, but the core and blue line wings appear relatively pristine.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ mag source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with a projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.

@@

We conducted orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}, but without a joint Gaussian Process modeling procedure \citep{2023ApJ...950..162T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\ms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  The period and $T_0$ were fixed from the \emph{TESS} transits, and we assume a circular orbit with $K>10$ \ms.  The semiamplitude prior acts as a mass constraint, excluding planets with masses so low that the observed radius would exceed the Hill radius (Roche lobe overflow).  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \ms, a relatively weak constraint, but consistent with broadly Saturn-mass planets.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  The semiamplitude prior accounts for the apparent improvement in the mass constraint.  Hypothetically, a truly Roche lobe overflow planet could be consistent with the available data, so our mass constraint could be considered an upper limit.

@@

The sky fiber and target fiber have slightly different throughputs and illumination properties, with the target fiber receiving $93\%$ of the flux of the sky reference fiber on average.  This ratio depends on wavelength and season at the few percent level.  We quantified the wavelength dependence by acquiring calibration observations of blank sky in both the target and sky fiber.  We applied this wavelength-based scale factor to each target spectrum's associated reference sky fiber to achieve near-imperceptible sky-line subtraction residuals.  A few lines exhibit residuals that may arise from genuine differences in the local atmospheric conditions between the target and sky fiber.

We masked spectral regions predicted to have significant telluric lines with a template generated by \texttt{telfit} \citep{2014AJ....148...53G} then shifted the spectral coordinates to their common barycentric-corrected reference frame \citep{2014PASP..126..838W}, as implemented in \texttt{astropy} \citep{2013A&A...558A..33A,2018AJ....156..123A,2022ApJ...935..167A}.  The spectral continuum was flattened and normalized to two pre-selected continuum indices highlighted as vertical blue bands in Figure \ref{fig:HPFheliumOverview}.  The sky-subtraction, telluric masking, barycentric correction, flattening, and all of our other standard pre-processing steps were carried out in the open-source Python interface \texttt{muler} \citep{2022JOSS....7.4302G}.

@@

>>>>

@@

Inflated hot Saturns stand out as an especially extreme category that should exhibit mass loss.  This category is defined as having masses comparable to Saturn's ($M_p \sim 0.3 M_\mathrm{Jup}$), with equilibrium temperatures high enough to expect radius inflation ($T_\mathrm{eq}>1000\;$K).  Their low gravitational potentials should let go of their atmospheres more readily than their hot Jupiter counterparts.  Their lower gravity implies larger atmospheric scale heights, making them easier to detect in transmission spectroscopy.  And their large transit depths and short periods should make them readily detectable in transit searches in large numbers, like hot Jupiters.

@@

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Here, lightly thermally ionized atmospheric flows induce drag in a planetary magnetic field, weakly coupling the stellar incident energy into the planetary interior.  A key prediction of Ohmic dissipation is that the anomalous heating efficiency $\epsilon(T_\mathrm{eq})$ should exhibit a peak around $T_\mathrm{eq}\sim1500-2000\;$K \citep{2012ApJ...745..138M,2014ApJ...794..132R,2016ApJ...819..116G}. \citet{2018AJ....155..214T} favored Ohmic dissipation as the mechanism responsible for inflating hot Jupiters by showing that the observed sample of inflated planets implies an anomalous heating efficiency peak at equilibrium temperatures of $\sim1500\;$K.  Therefore Ohmic dissipation stands out as a leading physical mechanism for inflation and mass loss.

@@

A runaway inflation scenario predicts an inevitable and profound mass loss rate for inflated hot Saturns.  As the planet's atmosphere overflows its Roche lobe at an ever-increasing pace, instantaneous mass loss rates may exceed $\dot{M}>10^{13}$ g/s, over 10$\times$ larger than those previously seen in atmospheric escape measurements to date \citep{2022arXiv221116243D}.  Inflated hot Saturns, therefore, make excellent targets for direct measurement of atmospheric escape and for testing the underlying mechanisms of mass loss.

@@

The evolutionary state of HAT-P-67~b offers even more intriguing possibilities.  The planet may have undergone re-inflation \citep{2022AJ....163...53S, 2022AJ....163..120G, 2023arXiv230306728G}, as the host star evolved through the subgiant phase-- with the planet's insolation increasing with the star's rapid luminosity jump in this part of the short-lived HR diagram.  The tidal gravity of the nearby massive ($M_\star \sim1.6\;M_\odot$) host star may amplify mass loss rates even further \citep{2007A&A...472..329E, 2023ApJ...945L..36T}. Its status as a rare inflated hot Saturn makes HAT-P-67~b a promising laboratory, uniquely suited for testing the runaway inflationary predictions of Ohmic dissipation and XUV photoevaporation.

@@

The observation schedule was strategically restricted to the spring season when the Barycentric Earth Radial Velocity (BERV) would Doppler shift telluric absorption lines sufficiently far away from the core of the Helium 10833 \AA~ feature \citep{2022ApJ...939L..11S}.  The small BERV still means the redward Helium line wing has significant telluric contamination between 10834$-$10836 \AA, but the core and blue line wings appear relatively pristine.  Sky emission lines land in this region, but are more easily mitigated by HPF's simultaneous sky reference fiber.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}, well-separated from the planet-host star and not a source of contamination for the HPF observations.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ mag source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with a projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.

@@

We conducted orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}, but without a joint Gaussian Process modeling procedure \citep{2023ApJ...950..162T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\ms$, a few times larger than the typical Keck HIRES measurements, due to the lower information content in the near-IR than the visible for this relatively rapidly rotating F star.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued.  The period and $T_0$ were fixed from the \emph{TESS} transits, and we assume a circular orbit with $K>10$ \ms.  The semiamplitude prior acts as a mass constraint, excluding planets with masses so low that the observed radius would exceed the Hill radius (Roche lobe overflow).  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \ms, a relatively weak constraint, but consistent with broadly Saturn-mass planets.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  The semiamplitude prior accounts for the apparent improvement in the mass constraint.  Hypothetically, a truly Roche lobe overflow planet could be consistent with the available data, so our mass constraint could be considered an upper limit.

@@

The sky fiber and target fiber have slightly different throughputs and illumination properties, with the target fiber receiving $93\%$ of the flux of the sky reference fiber on average.  This ratio depends on wavelength and season at the few percent level.  We quantified the wavelength dependence by acquiring calibration observations of blank sky in both the target and sky fiber.  We applied this wavelength-based scale factor to each target spectrum's associated reference sky fiber to achieve sky-line subtraction residuals typically less than the photon noise \citep{2022JOSS....7.4302G}.  A few lines exhibit residuals that may arise from genuine differences in the local atmospheric conditions between the target and sky fiber.

The HET's fixed-altitude design means that the airmass remains relatively constant across all pointings, lowering the telluric line variability compared to fully steerable telescopes which may sample a wider range in airmasses.  Variability in atmospheric conditions still makes it difficult to mitigate telluric lines to within the photon noise limit, so we masked spectral regions predicted to have significant telluric lines with a template generated by \texttt{telfit} \citep{2014AJ....148...53G}.  We then shifted the spectral coordinates to their common barycentric-corrected reference frame \citep{2014PASP..126..838W}, as implemented in \texttt{astropy} \citep{2013A&A...558A..33A,2018AJ....156..123A,2022ApJ...935..167A}.  The spectral continuum was flattened and normalized to two pre-selected continuum indices highlighted as vertical blue bands in Figure \ref{fig:HPFheliumOverview}.  The sky-subtraction, telluric masking, barycentric correction, flattening, and all of our other standard pre-processing steps were carried out in the open-source Python interface \texttt{muler} \citep{2022JOSS....7.4302G}.

@@

    The Center for Exoplanets and Habitable Worlds is supported by the Pennsylvania State University and the Eberly College of Science.
